R tools for accessing research literature for text mining

### 2019-08-23 manuscript.Rmd
Typo
fix #3 text-mining to text minig

---
layout: review, 11pt
linenumbers: true
title: "R tools for accessing research literature for text mining"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: myrmecocystus(at)gmail.com
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
abstract: |
  Text-mining is a powerful method for answering research questions. However,
  getting texts to extract information can be a daunting and complicated task.
  The primary reason for this is publishers. There are thousands of different
  publishers, each with their own licenses, URL patterns, access options, and
  more. Layered on top of that is the varied access each user has based on
  their institutional affiliation. Here, I introduce a suite of software packages
  in the R programming language for fetching texts. The tapestry of different
  publishers, access levels, and other factors requires a patchwork of approaches
  for getting texts to users. The flagship R package called fulltext attempts
  to simplify search and retrieval of texts for text-mining. The fulltext package,
  along with many others, make acquiring texts easier than ever, facilitating
  answering research questions with text-mining.
bibliography: components/references.bib
csl: components/peerj.csl
documentclass: components/elsarticle
output:
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true
---

@@

Text-mining is the broad term associated with pulling information
out of articles. Given the importance of text-mining, good
text-mining tools are needed to make it easier for researchers to do.
Graphical user interface (GUI) based text mining tools are available
[e.g., @ba2016interoperability; @Munoz2019; @Canada2017] and some research papers
have used them [@Chaix2019], but given the urgent recent call to action for
more reproducible research [@OSC2015; @Camerer2016; @Camerer2018],
we must move away from GUI based tools as fast as possible. A number of examples
of programmatic tools can be found in the literature. For example, @Sinclair2016
present a tool in Python called seqenv for the domain specific task of
linking sequences to environments through text mining.

Most recent text-mining papers do not use programmatic approaches, highlighting
the need for more programmatic text-mining tools, and increased
discussion of those tools to increase awareness. For example, many papers 
search Web of Science using their web interface, and downloading papers manually 
[@Ding2018; @McCallen2019]. Many of these papers doing GUI based searching and 
paper downloads are using R or Python downstream for analysis; replacing 
GUI based data acquisition with programmatic approaches will improve 
research.

@@

Here, I present an overview of text-mining tools in the R programming
language, not for text-mining analysis, but rather those
tools for searching for, acquiring, and extracting parts of texts
(e.g., title, abstract, authors). Most of the packages presented here are
part of the rOpenSci suite (https://ropensci.org/).

@@

Of the non-machine readable types, the most noteable is the Portable
Document Format (PDF)[^pdf]. These can be broken out into two groups: text based
PDFs and scanned PDFs. The former are converted from digital versions of
various kinds (MS Word, OpenOffice, LaTeX, markdown, etc.), while the
latter are created by scanning print articles to a PDF format. Text-based
PDFs are much better for text-mining purposes as plain text can be extracted
easily in R with [pdftools][], a binding to [libpoppler][]. However, with
scanned PDFs, text must be extracted using Optimal Character Recognition
(OCR; see R package [tesseract][]), which isn't always a clean solution,
especially compared to true text based PDFs.

The reality in scholarly publishing is all publishers, if they provide
any access to their articles, only provide PDF format. Very few publishers,
with some quite large (Elsevier, Pensoft, PLOS), provide XML format. Although
most publishers most likely have the XML behind each of their articles,
they for some indefensible reason do not share it - making text-mining more
difficult. Some provide plain text (Elsevier). I only know of one publisher
that provides full text as JSON (PLOS). Thus, text-mining, in most cases,
will require extracting text from PDFs.

@@

On top of the vast array of different data sources is the varied ways that 
search is implemented in each source. Most sources are probably using Solr
or Elasticsearch under the hood, though we can't know this for sure as most 
do not make their software infrastructure public knowledge. Nonetheless, data
sources differ in how search works from the user perspective. For example, some 
provide wild card search and some do not. Some sources are searching full text
of articles, while others only search metadata (i.e., title, authors, abstract).
In addition, each source has a different set of metadata/full text available.
In brief, the same search against different sources produces different results.
Some text-mining research articles perform the same search against many
different sources (refs), while others choose just one source.

@@

One can search by DOI or search citation data to get citations. In addition, Crossref has a
text-mining opt-in program for publishers. The result of this is that some
publishers provide URLs for full text content of their articles. The majority of these links
are pay-walled, while some are open access. Using any of the various tools for working with
Crossref data, you can filter your search to get only articles with full text links, and further
to get only articles with full text links that are open access.

>>>>

---
layout: review, 11pt
linenumbers: true
title: "R tools for accessing research literature for text mining"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: myrmecocystus(at)gmail.com
    footnote: Corresponding author
address:
  - code: cstar
    address: |
      rOpenSci, Museum of Paleontology, University of California, Berkeley, CA, USA
abstract: |
  Text mining is a powerful method for answering research questions. However,
  getting texts to extract information can be a daunting and complicated task.
  The primary reason for this is publishers. There are thousands of different
  publishers, each with their own licenses, URL patterns, access options, and
  more. Layered on top of that is the varied access each user has based on
  their institutional affiliation. Here, I introduce a suite of software packages
  in the R programming language for fetching texts. The tapestry of different
  publishers, access levels, and other factors requires a patchwork of approaches
  for getting texts to users. The flagship R package called fulltext attempts
  to simplify search and retrieval of texts for text mining. The fulltext package,
  along with many others, make acquiring texts easier than ever, facilitating
  answering research questions with text mining.
bibliography: components/references.bib
csl: components/peerj.csl
documentclass: components/elsarticle
output:
  pdf_document:
    template: components/elsarticle.latex
    keep_tex: true
    fig_caption: true
---

@@

Text mining is the broad term associated with pulling information
out of articles. Given the importance of text mining, good
text mining tools are needed to make it easier for researchers to do.
Graphical user interface (GUI) based text mining tools are available
[e.g., @ba2016interoperability; @Munoz2019; @Canada2017] and some research papers
have used them [@Chaix2019], but given the urgent recent call to action for
more reproducible research [@OSC2015; @Camerer2016; @Camerer2018],
we must move away from GUI based tools as fast as possible. A number of examples
of programmatic tools can be found in the literature. For example, @Sinclair2016
present a tool in Python called seqenv for the domain specific task of
linking sequences to environments through text mining.

Most recent text mining papers do not use programmatic approaches, highlighting
the need for more programmatic text mining tools, and increased
discussion of those tools to increase awareness. For example, many papers 
search Web of Science using their web interface, and downloading papers manually 
[@Ding2018; @McCallen2019]. Many of these papers doing GUI based searching and 
paper downloads are using R or Python downstream for analysis; replacing 
GUI based data acquisition with programmatic approaches will improve 
research.

@@

Here, I present an overview of text mining tools in the R programming
language, not for text mining analysis, but rather those
tools for searching for, acquiring, and extracting parts of texts
(e.g., title, abstract, authors). Most of the packages presented here are
part of the rOpenSci suite (https://ropensci.org/).

@@

Of the non-machine readable types, the most noteable is the Portable
Document Format (PDF)[^pdf]. These can be broken out into two groups: text based
PDFs and scanned PDFs. The former are converted from digital versions of
various kinds (MS Word, OpenOffice, LaTeX, markdown, etc.), while the
latter are created by scanning print articles to a PDF format. Text-based
PDFs are much better for text mining purposes as plain text can be extracted
easily in R with [pdftools][], a binding to [libpoppler][]. However, with
scanned PDFs, text must be extracted using Optimal Character Recognition
(OCR; see R package [tesseract][]), which isn't always a clean solution,
especially compared to true text based PDFs.

The reality in scholarly publishing is all publishers, if they provide
any access to their articles, only provide PDF format. Very few publishers,
with some quite large (Elsevier, Pensoft, PLOS), provide XML format. Although
most publishers most likely have the XML behind each of their articles,
they for some indefensible reason do not share it - making text mining more
difficult. Some provide plain text (Elsevier). I only know of one publisher
that provides full text as JSON (PLOS). Thus, text mining, in most cases,
will require extracting text from PDFs.

@@

On top of the vast array of different data sources is the varied ways that 
search is implemented in each source. Most sources are probably using Solr
or Elasticsearch under the hood, though we can't know this for sure as most 
do not make their software infrastructure public knowledge. Nonetheless, data
sources differ in how search works from the user perspective. For example, some 
provide wild card search and some do not. Some sources are searching full text
of articles, while others only search metadata (i.e., title, authors, abstract).
In addition, each source has a different set of metadata/full text available.
In brief, the same search against different sources produces different results.
Some text mining research articles perform the same search against many
different sources (refs), while others choose just one source.

@@

One can search by DOI or search citation data to get citations. In addition, Crossref has a
text mining opt-in program for publishers. The result of this is that some
publishers provide URLs for full text content of their articles. The majority of these links
are pay-walled, while some are open access. Using any of the various tools for working with
Crossref data, you can filter your search to get only articles with full text links, and further
to get only articles with full text links that are open access.

### 2019-08-23 manuscript.tex
fix #3 text-mining to text minig

  \begin{abstract}
  Text-mining is a powerful method for answering research questions.
  However, getting texts to extract information can be a daunting and
  complicated task. The primary reason for this is publishers. There are
  thousands of different publishers, each with their own licenses, URL
  patterns, access options, and more. Layered on top of that is the varied
  access each user has based on their institutional affiliation. Here, I
  introduce a suite of software packages in the R programming language for
  fetching texts. The tapestry of different publishers, access levels, and
  other factors requires a patchwork of approaches for getting texts to
  users. The flagship R package called fulltext attempts to simplify
  search and retrieval of texts for text-mining. The fulltext package,
  along with many others, make acquiring texts easier than ever,
  facilitating answering research questions with text-mining.
  \end{abstract}

@@

Text-mining is the broad term associated with pulling information out of
articles. Given the importance of text-mining, good text-mining tools
are needed to make it easier for researchers to do. Graphical user
interface (GUI) based text mining tools are available (e.g., Ba \&
Bossy, 2016; Cañada et al., 2017; Muñoz, Kissling \& Loon, 2019) and
some research papers have used them (Chaix et al., 2019), but given the
urgent recent call to action for more reproducible research (Open
Science Collaboration, 2015; Camerer et al., 2016, 2018), we must move
away from GUI based tools as fast as possible. A number of examples of
programmatic tools can be found in the literature. For example, Sinclair
et al. (2016) present a tool in Python called seqenv for the domain
specific task of linking sequences to environments through text mining.

Most recent text-mining papers do not use programmatic approaches,
highlighting the need for more programmatic text-mining tools, and
increased discussion of those tools to increase awareness. For example,
many papers search Web of Science using their web interface, and
downloading papers manually (Ding, Li \& Fan, 2018; McCallen et al.,
2019). Many of these papers doing GUI based searching and paper
downloads are using R or Python downstream for analysis; replacing GUI
based data acquisition with programmatic approaches will improve
research.

@@

Here, I present an overview of text-mining tools in the R programming
language, not for text-mining analysis, but rather those tools for
searching for, acquiring, and extracting parts of texts (e.g., title,
abstract, authors). Most of the packages presented here are part of the
rOpenSci suite (\url{https://ropensci.org/}).

@@

Of the non-machine readable types, the most noteable is the Portable
Document Format (PDF)\footnote{\url{https://en.wikipedia.org/wiki/PDF}}.
These can be broken out into two groups: text based PDFs and scanned
PDFs. The former are converted from digital versions of various kinds
(MS Word, OpenOffice, LaTeX, markdown, etc.), while the latter are
created by scanning print articles to a PDF format. Text-based PDFs are
much better for text-mining purposes as plain text can be extracted
easily in R with \href{https://github.com/ropensci/pdftools}{pdftools},
a binding to \href{https://poppler.freedesktop.org/}{libpoppler}.
However, with scanned PDFs, text must be extracted using Optimal
Character Recognition (OCR; see R package
\href{https://github.com/ropensci/tesseract}{tesseract}), which isn't
always a clean solution, especially compared to true text based PDFs.

The reality in scholarly publishing is all publishers, if they provide
any access to their articles, only provide PDF format. Very few
publishers, with some quite large (Elsevier, Pensoft, PLOS), provide XML
format. Although most publishers most likely have the XML behind each of
their articles, they for some indefensible reason do not share it -
making text-mining more difficult. Some provide plain text (Elsevier). I
only know of one publisher that provides full text as JSON (PLOS). Thus,
text-mining, in most cases, will require extracting text from PDFs.

@@

Some may not realize access to articles varies with IP address so that
access from campus vs.~from home (if not on a VPN) will drastically
differ. Sometimes a VPN is required, and this can provide a significant
technical hurdle to users attempting to do text-mining work.

One final hurdle in text-mining comes unsurprisingly from Elsevier. They
use so-called ``fences'' for programmatic access. That is, even if a
person trying to get an article programmatically their institution has
access to and they have access to, and they are on the correct IP
address, they may still not get access to an Elsevier article. Elsevier
puts in place these fences and only if you contact their technical team
directly can you get these fences removed, and only then on a per
institution basis.

@@

\href{https://github.com/ropensci/fulltext}{fulltext} is a general
purpose R package for the data part of text-mining: search for articles,
get links to articles, get article abstracts, and fetch full text of
articles. The \texttt{fulltext} package is always adding additional data
sources as time allows (See Table 1). Starting from searching for
articles, the outputs of search can be fed into a function to get links
to those articles, or to get abstracts for those articles, or to fetch
their full text. The following is a breakdown of the major distinct
parts of \texttt{fulltext}.

>>>>

  \begin{abstract}
  Text mining is a powerful method for answering research questions.
  However, getting texts to extract information can be a daunting and
  complicated task. The primary reason for this is publishers. There are
  thousands of different publishers, each with their own licenses, URL
  patterns, access options, and more. Layered on top of that is the varied
  access each user has based on their institutional affiliation. Here, I
  introduce a suite of software packages in the R programming language for
  fetching texts. The tapestry of different publishers, access levels, and
  other factors requires a patchwork of approaches for getting texts to
  users. The flagship R package called fulltext attempts to simplify
  search and retrieval of texts for text mining. The fulltext package,
  along with many others, make acquiring texts easier than ever,
  facilitating answering research questions with text mining.
  \end{abstract}

@@

Text mining is the broad term associated with pulling information out of
articles. Given the importance of text mining, good text mining tools
are needed to make it easier for researchers to do. Graphical user
interface (GUI) based text mining tools are available (e.g., Ba \&
Bossy, 2016; Cañada et al., 2017; Muñoz, Kissling \& Loon, 2019) and
some research papers have used them (Chaix et al., 2019), but given the
urgent recent call to action for more reproducible research (Open
Science Collaboration, 2015; Camerer et al., 2016, 2018), we must move
away from GUI based tools as fast as possible. A number of examples of
programmatic tools can be found in the literature. For example, Sinclair
et al. (2016) present a tool in Python called seqenv for the domain
specific task of linking sequences to environments through text mining.

Most recent text mining papers do not use programmatic approaches,
highlighting the need for more programmatic text mining tools, and
increased discussion of those tools to increase awareness. For example,
many papers search Web of Science using their web interface, and
downloading papers manually (Ding, Li \& Fan, 2018; McCallen et al.,
2019). Many of these papers doing GUI based searching and paper
downloads are using R or Python downstream for analysis; replacing GUI
based data acquisition with programmatic approaches will improve
research.

@@

Here, I present an overview of text mining tools in the R programming
language, not for text mining analysis, but rather those tools for
searching for, acquiring, and extracting parts of texts (e.g., title,
abstract, authors). Most of the packages presented here are part of the
rOpenSci suite (\url{https://ropensci.org/}).

@@

Of the non-machine readable types, the most noteable is the Portable
Document Format (PDF)\footnote{\url{https://en.wikipedia.org/wiki/PDF}}.
These can be broken out into two groups: text based PDFs and scanned
PDFs. The former are converted from digital versions of various kinds
(MS Word, OpenOffice, LaTeX, markdown, etc.), while the latter are
created by scanning print articles to a PDF format. Text-based PDFs are
much better for text mining purposes as plain text can be extracted
easily in R with \href{https://github.com/ropensci/pdftools}{pdftools},
a binding to \href{https://poppler.freedesktop.org/}{libpoppler}.
However, with scanned PDFs, text must be extracted using Optimal
Character Recognition (OCR; see R package
\href{https://github.com/ropensci/tesseract}{tesseract}), which isn't
always a clean solution, especially compared to true text based PDFs.

The reality in scholarly publishing is all publishers, if they provide
any access to their articles, only provide PDF format. Very few
publishers, with some quite large (Elsevier, Pensoft, PLOS), provide XML
format. Although most publishers most likely have the XML behind each of
their articles, they for some indefensible reason do not share it -
making text mining more difficult. Some provide plain text (Elsevier). I
only know of one publisher that provides full text as JSON (PLOS). Thus,
text mining, in most cases, will require extracting text from PDFs.

@@

Some may not realize access to articles varies with IP address so that
access from campus vs.~from home (if not on a VPN) will drastically
differ. Sometimes a VPN is required, and this can provide a significant
technical hurdle to users attempting to do text mining work.

One final hurdle in text mining comes unsurprisingly from Elsevier. They
use so-called ``fences'' for programmatic access. That is, even if a
person trying to get an article programmatically their institution has
access to and they have access to, and they are on the correct IP
address, they may still not get access to an Elsevier article. Elsevier
puts in place these fences and only if you contact their technical team
directly can you get these fences removed, and only then on a per
institution basis.

@@

\href{https://github.com/ropensci/fulltext}{fulltext} is a general
purpose R package for the data part of text mining: search for articles,
get links to articles, get article abstracts, and fetch full text of
articles. The \texttt{fulltext} package is always adding additional data
sources as time allows (See Table 1). Starting from searching for
articles, the outputs of search can be fed into a function to get links
to those articles, or to get abstracts for those articles, or to fetch
their full text. The following is a breakdown of the major distinct
parts of \texttt{fulltext}.
